---
title: "Likehood Estimates for Outcome Rank Given Correlated Variable"
author: "Rich Seiter"
date: "Sunday, July 27, 2014"
output: html_document
---

Based on idea described in my comment at http://infoproc.blogspot.com/2014/07/success-ability-and-all-that.html

Given a contributing factor X and a related outcome variable Y (with correlation r).  Calculate likelihood that a given X will have the largest Y.  An example case might be height as X and basketball ability as Y.  Here we will focus on abstract normal variables with mean 0 and standard deviation 1.  I am assuming a correlation of 0.65, but it will be interesting to vary that.

Ideally this would be computed exactly.  This analysis uses a simulation instead.

First we generate a single set of data to give a sense of what we will see.

```{r example}
Sys.time()
# These variables define our distribution
Xmean <- 0
Ymean <- 0
Xsd <- 1
Ysd <- 1
XYcor <- 0.65
#XYcor <- 0.4

# Derived variables
XYmean <- c(Xmean, Ymean)
XYcov <- matrix(c(1, XYcor, XYcor, 1), nrow=2, ncol=2)

# Create our multivariate normal distribution using mvrnorm in MASS
require(MASS)
 
#make this reproducible
set.seed(2)
 
# Don't make sample too large here, car scatterplot can be slow
myDraws <- mvrnorm(1e4, mu=XYmean, Sigma=XYcov)
myDraws <- data.frame(myDraws)
colnames(myDraws) <- c("X", "Y")

#check the mean, sd, and cor
apply(myDraws, 2, mean)
apply(myDraws, 2, sd)
cor(myDraws)

# Plot the distribution
require(car)
scatterplot(Y ~ X, data=myDraws, ellipse=TRUE, levels=c(0.5, 0.95, 0.99, 0.999, 0.9999))
```

Simulate Multiple Populations
-----------------------------

```{r simulation}
Sys.time()
# These values take about 15 minutes
# numRuns <- 1e4
# popSize <- 1e5
numRuns <- 1e4
popSize <- 1e5

# Quick version for debugging
# numRuns <- 1e2
# popSize <- 1e4

# Also record X for a top portion of Y
topProp <- 0.001 # Try 0.1%
topN <- popSize*topProp

XOfLargestY <- 1:numRuns
XOfTopPropY <- 1:(numRuns*topN)

#make this reproducible
set.seed(2)
for (i in 1:numRuns) {
  myDraws <- mvrnorm(popSize, mu=XYmean, Sigma=XYcov)
  myDraws <- data.frame(myDraws)
  colnames(myDraws) <- c("X", "Y")
  XOfLargestY[i] <- myDraws$X[which.max(myDraws$Y)]
  XOfTopPropY[((i-1)*topN+1):(i*topN)] <- myDraws$X[head(order(myDraws$Y, decreasing=TRUE), n=topN)]
}
Sys.time()
```

Look at the Results
-------------------

How often does the largest Y have a given X value?

```{r}
require(lattice)
densityplot(XOfLargestY)

# Take a closer look at this distribution
summary(XOfLargestY)
mean(XOfLargestY)
sd(XOfLargestY)
```

How likely is it that any given person at level X will have the best Y?

I'm not sure about these calculations (especially the attempt to normalize to a probability).  Feedback welcomed.

```{r}
obs_dvals <- density(XOfLargestY)
dvals <- dnorm(obs_dvals$x,mean=0,sd=1)
prob <- data.frame(x = obs_dvals$x,
                    y = obs_dvals$y / dvals / popSize # Attempt to normalize this to prob.
                    )

xyplot(y ~ x, prob, type="l")
xyplot(log10(y) ~ x, prob, type="l")
```

I believe the somewhat anomalous results at the extremes are due to sampling error.  For example, when I increased the popSize from 1e4 to 1e5 (numRuns = 1e2) the peak and dip at the right moved right by about 0.7 to correspond.  Then when I increased popSize to 1e6 the peak moved another ~0.3 right.  The center of the XOfLargestY distribution also shifted right and the probabilities for 3-4SD X decreased.

Now Look at the Top `r topProp*100`% of Y
-----------------------------------------

How often does the top `r topProp*100`% of Y have a given X value?

```{r}
densityplot(XOfTopPropY)

# Take a closer look at this distribution
summary(XOfTopPropY)
mean(XOfTopPropY)
sd(XOfTopPropY)
```

How likely is it that any given person at level X will be in the top `r topProp*100`% of Y?

```{r}
obs_dvals1 <- density(XOfTopPropY)
dvals1 <- dnorm(obs_dvals1$x,mean=0,sd=1)
prob1 <- data.frame(x = obs_dvals1$x,
                    y = obs_dvals1$y / dvals1 / popSize * topN # Attempt to normalize this to prob.
                    )

xyplot(y ~ x, prob1, type="l")
xyplot(log10(y) ~ x, prob1, type="l")
```

Plot the densities together.

```{r}
plot(density(XOfLargestY), xlim=c(min(XOfTopPropY), max(XOfTopPropY)), col="red",
#plot(density(XOfLargestY), xlim=c(min(XOfLargestY), max(XOfLargestY)), col="red",
     main=paste("Compare Densities of Largest and Top", topProp*100, "% of Y"),
     xlab="X")
lines(density(XOfTopPropY))
legend("topright", legend=c("Largest", paste("Top", topProp*100, "%")), lty=1, col=c("red", "black"))
```

Note limiting of X axis.  I think this is the valid range for inference.

```{r}
#plot(prob$x, log10(prob$y), xlim=c(min(XOfTopPropY), max(XOfTopPropY)),
plot(prob$x, log10(prob$y), xlim=c(min(XOfLargestY), max(XOfLargestY)),
     ylim=c(min(log10(prob1$y)),max(log10(prob1$y))),
     col="red",
     main=paste("Compare Per-Person Likelihood of Largest and Top", topProp*100, "% of Y"),
     xlab="X", ylab="log10(probability)", type="l")
lines(prob1$x, log10(prob1$y))
legend("topleft", legend=c("Largest", paste("Top", topProp*100, "%")), lty=1, col=c("red", "black"))
```

Comments
--------

Clearly this simulation model has many assumptions.  The two that particularly concern me are:  
1. Normality (e.g. IQ is believed to have fat tails)  
2. That the sample population is large enough.  I saw consistent changes when increasing popSize.  I finally chose 1e5 as a compromise.

I think it best to only really consider the results in the X range where XOfLargestY has at least one value (and possibly ignore apparent extreme outliers).


Other Notes
-----------

A relevant point by another commenter: [True or False: Half of All 7-Footers are in the NBA](http://www.truthaboutit.net/2012/05/true-or-false-half-of-all-7-footers-are-in-the-nba.html)