---
title: "Terman Study Thoughts"
author: "Rich Seiter"
date: "Monday, July 28, 2014"
output: html_document
---

Emil Kirkegaard posted mean and SD for participants in the Terman study in a comment to http://infoproc.blogspot.com/2014/07/success-ability-and-all-that.html  
"The average Termite was 152, sd=15."

This seemed high if IQ is a normal distribution and Terman Study participants can be just viewed as a truncated sample of it over 135 so I decided to simulate that and see how the numbers compared.  This was apparently a mistake.  Changed cutoff to 140.

```{r}
cutoff <- 140

#make this reproducible
set.seed(2)
iq <- rnorm(1e6, mean=100, sd=15)
summary(iq)
studyIQ <- iq[iq>cutoff]
summary(studyIQ)
mean(studyIQ)
sd(studyIQ)

plot(density(studyIQ), xlab="IQ", main="Simulated IQ Density for Terman Study")
abline(v=mean(studyIQ))
```

Let's try a lognormal distribution.  See http://en.wikipedia.org/wiki/Log-normal_distribution for details of how this is parameterized.  I'm not sure I have the right distribution...

```{r}
m <- 100
v <- 15^2
ml <- log(m^2 / (v + m^2))
sl <- sqrt(log(1+v/m^2))
#make this reproducible
set.seed(2)
iql <- m * rlnorm(1e6, meanlog=ml, sdlog=sl)
mean(iql)
sd(iql)
summary(iql)
studyIQL <- iql[iql>cutoff]
summary(studyIQL)
mean(studyIQL)
sd(studyIQL)

plot(density(studyIQL), xlab="IQ", main="Simulated Lognormal IQ Density for Terman Study")
abline(v=mean(studyIQL))
```

This [link](http://www.davidsongifted.org/db/Articles_id_10192.aspx) states that 26/1500 of Terman's subjects had IQs > 180. (250,000 screened?)

May want to see [Children Above 180 IQ](https://www.worldcat.org/title/children-above-180-iq-stanford-binet-origin-and-development/oclc/2680258&referer=brief_results)  Also see [Hathitrust](http://catalog.hathitrust.org/Record/000382871) and [Google Books](http://books.google.com/books?id=h4J9AAAAMAAJ&dq=editions:OCLC313182987) and [Review](http://hiqnews.megafoundation.org/Children_Above_180_IQ.htm)

